---
output: html_document
---
## Analysis of environmental events that have the greatest impact in the US.
    Created by Luciano Guerra

### Synonpsis  

In this report we analyse the last years data recorded by the U.S. National Oceanic and Atmospheric Administration's (NOAA) in their storm database. In this database, we have all events from 1950 until the end of November 2011. In this database NOAA tracks characteristics of major storms and weather events in the United States.And the final goal of this analisys is to identify the most dangerous events in two ways: fatalities and injuries produced by one side, and property damages by the other side.

### Data Processing
All data is provided by  [U.S. National Oceanic and Atmospheric Administration's (NOAA)](https://www.noaa.gov)
Analisys data is downloaded from  [stormData](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Fstormdata.csv.bz2)
and is saved in  [working directory](E:/Luciano/R_WorkingDir/Reproducible Research/RepData_PeerAssessment2").

Prepare working folder:
```{r createwfolder, echo=TRUE, echo=TRUE}
# URLfile <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Fstormdata.csv.bz2"
mywd <- "E:/Luciano/R_WorkingDir/Reproducible Research/RepData_PeerAssessment2"
actwd <- getwd()
if (actwd != mywd){
    setwd(mywd)
}
rm(list = ls())
filelist <- dir()
if ("FStormData.csv.bz2" %in% filelist){
    print("Start analisys.")
} else{
    print("To perform analisys donwload file and copy it with Fstormdata.bz2 name")
    knitr::knit_exit()
}
```
Unzip data and trasnfer all information to a data frame
Load and clean data. Transform all relevant variables in uppercase and date is formated
```{r loaddata, echo=TRUE, cache=TRUE}
    stormdata <- read.csv(file = "./FStormData.csv.bz2")
    # Clean data
    stormdata$EVTYPE <- (toupper(stormdata$EVTYPE))
    stormdata$PROPDMGEXP <- toupper(stormdata$PROPDMGEXP)
    stormdata$CROPDMGEXP <- toupper(stormdata$CROPDMGEXP)
    stormdata$BGN_DATE <- as.POSIXlt(stormdata$BGN_DATE, format = "%m/%d/%Y %H:%M:%S")
    stormdata$PROPDMG.MTPL <- ifelse(stormdata$PROPDMGEXP %in% c('h','H'),100, 
                            ifelse(stormdata$PROPDMGEXP %in% c('k','K'),1000,
                          ifelse(stormdata$PROPDMGEXP %in% c('m','M'), 1e6,
                          ifelse(stormdata$PROPDMGEXP %in% c('b','B'),1e9,1) ) ))

    stormdata$CROPDMG.MTPL <- ifelse(stormdata$PROPDMGEXP %in% c('h','H'),100,
                              ifelse(stormdata$CROPDMGEXP %in% c('k','K'),1000,
                              ifelse(stormdata$CROPDMGEXP %in% c('m','M'), 1e6,
                              ifelse(stormdata$CROPDMGEXP %in% c('b','B'),1e9,1) ) ))
    stormdata$NewPROPDMG <- stormdata$PROPDMG * stormdata$PROPDMG.MTPL
    stormdata$NewCROPDMG <- stormdata$CROPDMG * stormdata$CROPDMG.MTPL
```
#### Dangerous Events with respect to Population Health
Prepare data to read fatalities and injuries by type of event
```{r headlthdata, cache=TRUE, echo=TRUE}
    totalinjuries <- with(stormdata, aggregate(x = INJURIES, by = list(EVTYPE), FUN = sum))
    totalfatalities <- with(stormdata, aggregate(x = FATALITIES, by = list(EVTYPE), FUN = sum))
    names(totalinjuries) <- c("Event", "Injuries")
    names(totalfatalities) <- c("Event", "Fatalities")
    totalinjsort <- totalinjuries[order(totalinjuries$Injuries, decreasing = TRUE),]
    totalfatsort <- totalfatalities[order(totalfatalities$Fatalities, decreasing = TRUE),]
    totalinjsort[1:20,]
    totalfatsort[1:20,]
```

#### Economic Efects of Weather Events
Prepare data to read crop and properties damages
```{r damdata, cache=TRUE, echo=TRUE}
    totalcrop <- with(stormdata, aggregate(x = NewCROPDMG, by = list(EVTYPE), FUN = sum))
    totalprop <- with(stormdata, aggregate(x = NewPROPDMG, by = list(EVTYPE), FUN = sum))
    names(totalcrop) <- c("Event", "Cropdamage")
    names(totalprop) <- c("Event", "Propdamage")
    totalcropsort <- totalcrop[order(totalcrop$Cropdamage, decreasing = TRUE),]
    totalpropsort <- totalprop[order(totalprop$Propdamage, decreasing = TRUE),]
    totalcropsort[1:20,]
    totalpropsort[1:20,]
    totaldam <- merge(totalcrop, totalprop)
    totaldam$cropandpropdam <- totaldam$Cropdamage + totaldam$Propdamage
    totaldamsort <- totaldam[order(totaldam$cropandpropdam, decreasing = TRUE),]
```
The most harmful with respect to population health:

- Injuries: `r totalinjsort$Injuries[1]` 
- Fatalities: `r totalfatsort$Fatalities[1]` 

Types of events that have the greatest economic consequences:

- Crop damages: `r totalcropsort$Cropdamage[1]`
- Properties damages: `r totalpropsort$Propdamage[1]`

### Results
####  1.- Health impact

More severe weather events by number of injured.
```{r injuriesplot, echo=TRUE}
    par(mar = c(12,4,3,2), las = 3)
    barplot(totalinjsort$Injuries[1:20], names.arg = totalinjsort$Event[1:20], 
            col = rainbow (20, start=0, end=0.5), main = "More severe weather events by number of injured.",
            ylab = "Injuries")
```

More severe weather events by number of fatalities.

```{r fatalitiesplot, echo=TRUE}
    par(mar = c(12,4,3,2), las = 3)
    barplot(totalfatsort$Fatalities[1:20], names.arg = totalfatsort$Event[1:20], 
            col = rainbow (20, start=0, end=0.5), main = "More severe weather events by number of fatalities.",
            ylab = "Fatalities")
```

####  2.- Economic impact

Total economic impact of ordered by event

```{r cropandpropdam, echo=TRUE}
    par(mar = c(12,4,3,2), las = 3)
    barplot(totaldamsort$cropandpropdam[1:20]/1e9, names.arg = totaldamsort$Event[1:20], 
            col = rainbow (20, start=0, end=0.5), main = "Total economic impact of ordered by event.",
            ylab = "Total damages (billion $)")
```